# (PART) Geostatistical Data {-}

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE)
```

```{r, echo=FALSE}
library(ggplot2)
library(plotly)
library(gstat)
library(latex2exp)
library(tidyverse)
```


# Geostatistics

## Targeted Problems

Spatial data is considered as a random process $\{Z(s),s\in D\}$ in this part.   

- $D$: a is a fixed subset of $\mathbb{R}^d$ with positive d-dimensional volume.    
- $s$: locations varies _continuously_ throughout the region $D$.

Set `coalash` dataset as an example (Figure \@ref(fig:coalash-scatter)). $D$ is the region with values. and $s$ indicates percent of coalash in this location. Many kinds of exploratory statistics can be applied here to test stationarity, local stationarity and so on.

```{r, coalash-scatter, fig.cap = "Scatter Plot of CoalAsh Percent", echo = FALSE}
data("coalash")
plot_ly(x = coalash$x,  y = coalash$y,  z = coalash$coalash, 
        type='scatter3d',  mode='markers',  color = coalash$coalash,  alpha = 0.7) 
```

The key idea in this chapter is to model the above random process $\{Z(s),s\in D\}$ with values on known locations. Then inference of unknown locations can be made.

## Spatial Data Example

### Main Concepts

Variogram is the crucial parameter of geostatistics. 

> **Intrinsic Stationarity** \
Based on the random process mentioned above, intrinsic stationarity is defined through first differences
\begin{equation}
\begin{aligned}
&E\{Z(s+h) - Z(s)\} = 0,\\
&Var\{Z(s+h) - Z(s)\} = 2\gamma(h).
\end{aligned}
(\#eq:stationary2)
\end{equation}

> **Variogram** \
Variogram is $2\gamma(h)$ in the definition of intrinsic stationarity.

### Esitimation of Variogram

#### Estimator

The classical estimator of the variogram: \
\begin{equation}
2 \hat{\gamma}(h) \equiv \frac{1}{|N(h)|} \sum_{N(h)}\left(Z\left(s_{i}\right)-Z\left(s_{j}\right)\right)^{2}
(\#eq:v-estimation-classic)
\end{equation}

Estimator \@ref(eq:v-estimation-classic) is unbiased. However, it is badly affected by atypical observations due to square operation.

For an approximately symmetric distribution, there is a more robust approach to the estimation of the variogram by transforming the problem to location estimation. For a Gaussian process, we have:

\begin{equation}
2 \bar{\gamma}(\mathrm{h}) \equiv\left\{\frac{1}{|N(\mathrm{~h})|} \sum_{N(\mathrm{~h})}\left|Z\left(\mathrm{~s}_{i}\right)-Z\left(\mathrm{~s}_{j}\right)\right|^{1 / 2}\right\}^{4} /\left(0.457+\frac{0.494}{|N(\mathrm{~h})|}\right).
(\#eq:v-estimation-robust)
\end{equation}

Estimator \@ref(eq:v-estimation-robust) is also unbiased.

#### Visualization

**Variogram Cloud**: \
Cousin to estimator \@ref(eq:v-estimation-classic), variogram cloud is a useful diagnostic tool. Similarly, it can be badly affected by atypical observations. Variogram cloud in east-west direction of coalash data is shown in Figure \@ref(fig:v-cloud)

Here are steps to draw variogram cloud. \

- Group points with $y$ value;
- Calculate the classical estimator of the variogram $\frac{1}{|N(h)|}\cdot \displaystyle\sum_{N(h)}{(z(s_i) - z(s_j))^2}$. \
  - $N(h) = \{(s_i, s_j):\ s_i + he = s_j\}$
  - $e$: the unit vector of east direction (positive x axis)
  - Notice that y values of each pair $(s_i, s_j)$ are the same. But they may vary among pairs.
- Plot the corresponding boxplot. \
  - x axis: distance $h$
  - y axis: the classical estimator of the variogram with parameter h, Units are $(\% coal ash)^2$

```{r, v-cloud, fig.cap="Variogram Cloud in the East-West Direction", echo = FALSE}
get_df <- function(coalash, fun){
  df <- data.frame()
  for(j in min(coalash$y):max(coalash$y)){
    data <- coalash[which(coalash$y == j),]
    x_len <- length(data$x)
    mat1 <- matrix(rep(data$x, times = x_len), ncol = x_len)
    mat_h <- mat1 - t(mat1)
    
    mat2 <- matrix(rep(data$coalash, times = x_len), ncol = x_len)
    mat_z <- mat2 - t(mat2)
    df_current <- data.frame(h = as.numeric(mat_h), 
                             z = fun(as.numeric(mat_z))) %>% filter(h > 0)
    df <- rbind(df, df_current)
  }
  return(df)
}

df <- get_df(coalash, function(x){return(x**2)})
p <- ggplot(df, aes(x = as.factor(h), y = z)) + 
  geom_boxplot(aes(fill = h, color = h, alpha = 0.7), outlier.alpha = 0.7) +
  theme_light() +
  # labs(x = "h", y = TeX("$(Z(h+s)-Z(s))^2$"),
  labs(x = "h", y = ("Variogram"),
       title = "Variogram Cloud in the East-West Direction") +
  theme(plot.title = element_text(hjust = 0.5)) +
  guides(fill = "none", color = "none")
ggplotly(p)
```

**Square-Root-Differences Cloud**: \
The only different from variogram cloud: calculate the classical estimator of the variogram $\frac{1}{|N(h)|}\cdot \displaystyle\sum_{N(h)}{|Z(s_i) - Z(s_j)|^\frac12}$.

```{r, squre-cloud, fig.cap="Square-Root-Differences Cloud in the East-West Direction"}
df <- get_df(coalash, function(x){return(sqrt(abs(x)))})
p <- ggplot(df, aes(x = as.factor(h), y = z)) + 
  geom_boxplot(aes(fill = h, color = h, alpha = 0.7), outlier.alpha = 0.7) +
  theme_light() +
  labs(x = "h", y = ("Square-Root-Differences"),
       title = "Square-Root-Differences Cloud in the East-West Direction") +
  theme(plot.title = element_text(hjust = 0.5))
ggplotly(p)
```



**Pocket Plot**

The Pocket Plot tries to identify a localized area as being atypical with respect to a stationary mode. We plotting Pocket Plot \@ref(fig:pocket) in the North-South Direction follows the steps as below:  

- Get $\{|N_{jk}|\}$ matrix (symmetric).
  - Denote $|N_{jk}|$ as the value of row $j$ and column $k$ of the matrix
  - $N_{jk} = \{(s_j, s_k):\ s_j \in row_j, s_k \in row_k, x_{s_j} = x_{s_k}\}$
  - $|N_{jk}|$ means total pair number between row $j$ and $k$
- Get $\{Y_{jk}\}$ matrix (symmetric).
  - For there are 23 distinct value in y axis, the matrix has the dim of $23\times 23$
  - Denote $Y_{jk}$ as the value of row $j$ and column $k$ of the matrix, we have $Y_{jk} = \frac{1}{N_{jk}}\displaystyle\sum_{N_{jk}}|z(s_j) - z(s_k)|^{\frac12}$
- Calculate $Y_h$ vector.
  - $Y_h$ is the weighted mean of $Y_{jk}$ where $|j - k| = h$ 
- Calculate $P_{jk} = Y_{jk} - Y_h$
- Plot the boxplot.
  - x axis: index $j$
  - y axis: $P_{jk}$. 
  - There should be 23 points of each columns.

```{r}
get_Y_jk <- function(coalash, j, k){
  df_j <- coalash %>% filter(y == j) %>% select(x, coalash)
  df_k <- coalash %>% filter(y == k) %>% select(x, coalash)
  df <- merge(df_j, df_k, by = "x") %>%
    mutate(dif = sqrt(abs(coalash.x - coalash.y)))
  res <- mean(df$dif)
  num <- nrow(df)
  return(list(res, num))
}
```

```{r}
is_outlier <- function(x) {
  return(x < quantile(x, 0.25, na.rm = TRUE) - 1.5 * IQR(x, na.rm = TRUE) | x > quantile(x, 0.75, na.rm = TRUE) + 1.5 * IQR(x, na.rm = TRUE))
}
```

```{r}
# get Y_jk, N_jk matrix
row_len <- length(unique(coalash$y))
Y_jk <- matrix(nrow = row_len, ncol = row_len)
N_jk <- matrix(nrow = row_len, ncol = row_len)
for(i in 1:row_len){
  for(j in 1:row_len){
    jk_info <- get_Y_jk(coalash, i, j) # func `get_Y_jk` def hidden
    Y_jk[i,j] <- jk_info[[1]]
    N_jk[i,j] <- jk_info[[2]]
  }
}
colnames(Y_jk) <- 1:23
colnames(N_jk) <- 1:23
```

```{r}
# colnames(df_N_jk): "j", "k", "N_jk"
df_N_jk <- as.data.frame(as.table(N_jk)) %>%
  mutate_if(is.factor, as.integer) %>%
  rename(j = 1, k = 2, N_jk = 3)

# colnames(df_Y_jk): "j", "k", "Y_jk", "h", "N_jk"
df_Y_jk <- as.data.frame(as.table(Y_jk)) %>% 
  mutate_if(is.factor, as.integer)%>%
  transform(h = abs(Var1 - Var2)) %>%
  rename(j = 1, k = 2, Y_jk = 3) %>%
  left_join(df_N_jk, by = c('j', 'k')) %>%
  filter(h > 0)

# colnames(df_Y_h): "h", "N_h", "Y_h"
df_Y_h <- df_Y_jk %>% group_by(h) %>% 
  summarise(N_h = n(), Y_h = sum(Y_jk)/max(N_h))

# colnames(df_P_jk): "h", "j", "k", "Y_jk", "N_jk", "N_h", "Y_h", "P_jk", "outlier"
df_P_jk <- merge(df_Y_jk, df_Y_h, by = "h") %>%
  transform(P_jk = Y_jk - Y_h) %>%
  group_by(j) %>%
  mutate(outlier = ifelse(is_outlier(P_jk), k, as.numeric(NA)))
```

```{r, pocket, fig.cap="Pocket Plot in the North-South Direction"}
# Pocket plot
p <- ggplot(df_P_jk, aes(x = as.factor(j), y = P_jk)) +
  geom_boxplot(outlier.colour = 'red') +
  geom_text(aes(label = outlier), check_overlap = TRUE, nudge_x = 0.5) +
  theme_light() +
  labs(x = "row number", y = TeX("$P_{j\ k}$"),
  # labs(x = "row number", y = ("P_jk"),
       title = "Pocket Plot in the North-South Direction") +
  theme(plot.title = element_text(hjust = 0.5))
p
```


### Decomposition

The goal is to deompose data into large- and small-scale variation.

> **Large-Scale Variation**: The grand, column, and row effects.

> **Small-Scale Variation**: The residual.

- Like ANOVA, $F$ test within-rows and within-columns is first raised. However, it suffer from correlation among data points.   
- Another additive decomposition follows the idea of $data = all + row + column + residual$. Then for value of a location $(i,j)$ can be decomposed as $Y_{ij} = Y_{..} + (Y_{i.}-Y_{..}) + (Y_{.j}-Y_{..}) + (Y_{ij}-Y_{i.}-Y_{.j}+Y_{..})$, where a subscripted dot (·) denotes averaging over that subscript. This can be settled by _median polishing_ method.

**Median Polishing** \
Successively sweeps medians out of rows, then columns, then rows, then columns, and so on, accumulating them in "row," "column," and "all" registers, and leaves behind the table of residuals. 


## Stationary Process

The concept of _intrinsic stationarity_ has been introduced in section \@ref(main-concepts). Second-order (weak) stationarity is addressed in this section. It should be noticed that intrinsic stationarity is weaker than second-order stationarity.

> **Intrinsic Stationarity** \
Based on the random process mentioned above, intrinsic stationarity is defined through first differences
\begin{equation}
\begin{aligned}
&E\{Z(s+h) - Z(s)\} = 0,\\
&Var\{Z(s+h) - Z(s)\} = 2\gamma(h).
\end{aligned}
(\#eq:stationary2)
\end{equation}

> **Second-Order (Weak) Stationarity** \
When the random process meets the following assuptions, the process is second-order (weak) stationary. 
\begin{equation}
\begin{aligned}
&E(Z(s)) = 0, \ \forall s \in D \\
&Cov(Z(s_i), Z(s_j)) = C(s_i - s_j), \ \forall s_i, s_j \in D
\end{aligned}
(\#eq:stationary)
\end{equation}


Other important concepts related to stationary process as listed below. \

- **Covariogram**: Function $C(.)$ in \@ref(eq:stationary).
- **Isotropy**: When $C(s_i - s_j)$ only depends on $\Vert s_i - s_j\Vert$ but not the direction of $s_i - s_j$, then $C(.)$ is isotropic.
- **Anisotropy**: Contrary to isotropy, when $C(s_i - s_j)$ depends on both $\Vert s_i - s_j\Vert$ and the direction of $s_i - s_j$, then $C(.)$ is anisotropic.
    - Caused by underlying physical process evolving differentlly in space.
    - The Euclidean space is not appropriate for measuring distance between locations.
- **Ergodicity**: The property which allows expectations over the $\Omega \in \mathbb{R}^d}$ space to be estimated by spatial averages.
- **Strong Stationarity**: Strong stationarity refers to the property that the distribution of any finite sample of $Z(.)$ is independent of locations (translation invariant).


### Variogram & Covariogram {#vc}

Recall section \@ref(main-concepts), variogram is defined as $2\gamma(h) = Var(Z(s+h) - z(s)),\ \forall s, s+h\in D$. Some exploratory analysis of variogram has been seen in the example. Some related concepts are listed below.

- **Semiv-variogram**: $\gamma(h) = \frac12 Var(Z(s+h) - z(s))$.
- **Relative Variogram**: Variogram of transformed data. Sometimes, simple transformations lead to intrinsic stationary processes.
- **Nugget Effect**: $c_0$ is named as nugget effect, where $c_0 = \lim_{h\to 0}\gamma(h)$.   
    - When $h\to 0$, $Var(Z(s+h) - Z(s))\to Var(0)$, which should be exactly 0. This implies that nuggest effect should be 0 if assumptions are all met.
    - Causes of $c_0 > 0$: 
        - Measurement error;
        - Microscale variation.
    - Estimation: Extrapolating the variogram estimates fro lags closest to zero. for practically, it is impossible to calculate variogram with scale less than $\min \left\{\left\|s_{i}-s_{j}\right\|, 1 \leq i \leq j \leq n\right\}$.
- **Still**: The quantity $C(0)$ is called still. $C(0) - c_0$ is called the partial still.

Here are some other properties related to variogram and covariogram.

- Variogram is symmetric: $\gamma(h) = \gamma(-h),\ \gamma(0) = 0$.
- Speed of increase: $\lim _{|h| \rightarrow \infty} \frac{2 \gamma(h)}{|h|^{2}}=0$, which means that it is necessary for a variogram to increase slower than $|h|^2$.
- Under stationarity, the covariance function $C(h)$ is necessarily positive definite. 
- Under intrinsic stationarity, variograms must be conditionally negative-definite.

### Estimation of Variogram & Covariogram

#### Estimator of Variogram

Considering the definition of variogram, the most direct solution is moment estimator $$2 \hat{\gamma}(h) \equiv \frac{1}{|N(h)|} \sum_{N(h)}\left(Z\left(s_{i}\right)-Z\left(s_{j}\right)\right)^{2}$$ where $N(h) = \{(s_i, s_j): s_i - s_j = h\}$

Generally speaking, the variogram estimator can be expressed as a quadratic form \@ref(eq:qestimator)

\begin{equation}
2\hat \gamma (h) = Z'A(h)Z
(\#eq:qestimator)
\end{equation}

Denote $Var(Z)=\Sigma$, then the means and variance of $2\gamma(h)$ can be calculated as:

\begin{equation}
\begin{aligned} E(2 \hat{\gamma}(h)) &=\operatorname{trace}(A(h) \Sigma) \\ \operatorname{var}(2 \hat{\gamma}(h)) &=2 \operatorname{trace}(A(h) \Sigma A(h) \Sigma) \end{aligned}
(\#eq:matrixestimator)
\end{equation}

#### Estimator of Covariogram

For Covariogram, estimators are in line with basic ideas of variogram estimators. In the same way, the most direct thought of moment estimator is as Formula \@ref(eq:mcov-estimator)

\begin{equation}
\hat{C}(h) \equiv \frac{1}{|N(h)|} \displaystyle\sum_{N(b)}\left(Z\left(s_{i}\right)-\bar{Z}\right)\left(Z\left(s_{j}\right)-\bar{Z}\right)
\end{equation}

where $\bar{Z}=\sum_{i=1}^{n} Z\left(s_{i}\right) / n$.

### Validity

Not all functions can be variograms or covariogram. There are some prerequisites ensuring validity of them.

#### Valid Variogram

- Under intrinsic stationarity, variogram must be conditionally negative definite, which means \@ref(eq:valid-variogram-cnd)

\begin{equation}
\sum_{i, j} a_{i} a_{j} 2 \gamma\left(s_{i}-s_{j}\right) \leq 0$ for $\sum a_{i}=0
(\#eq:valid-variogram-cnd)
\end{equation}

- Continuous conditionally negative definite function corresponds to an intrinsically stationary stochastic process.
- Valid variogram forms a convex cone.
- Theorem \@ref(thm:valid-variogram) provides us with several ways to check variogram validity.


::: {.theorem #valid-variogram}
If $2 \gamma(.)$ is continuous on $\mathbb{R}^{d}$ and $\gamma(0)=0$, then the following are equivalent:

1. $2 \gamma(.)$ is conditionally negative definite; \
2. For all $a>0$, $\exp (-a \gamma(.))$ is positive definite. \
3. $2 \gamma(h)=Q(h)+\int \frac{1-\cos \left(\omega^{\prime} h\right)}{\|\omega\|^{2}} G(d \omega)$, where $Q(.) \geq 0$ is a quadratic form and $G(.)$ is a positive, symmetric measure continuous at the origin that $\int\left(1+\|\omega\|^{2}\right)^{-1} G(d \omega)<\infty$.
:::

#### Valid Covariogram

- Under second-order stationarity, the covariance function $C(h)$ is necessarily positive definite. That is, for any real numbers $a_1,...,a_n$ and locations $s_1,...,s_n$,

\beegin{equation}
\operatorname{Var}\left(\sum_{i=1}^{n} a_{i} Z\left(s_{i}\right)\right)=\sum_{i, j} a_{i} a_{j} C\left(s_{i}-s_{j}\right) \geq 0
(\#eq:valid-covariogram)
\end{equation}

- Spectral representation form
    - Accoding to Theorem \@ref(thm:valid-covariogram), a continuous function $C(h)$ is positive-definite if and only if it has a spectral representation $C(h)=\int \cos \left(\omega^{\prime} h\right) G(d \omega)$, where $G$ is a positive bounded symmetric measure.  
    - Spectral density and distribution
        -  $G/C(0)$ is the spectral distribution function
        - If $G(d \omega)=g(\omega) d \omega$, $g/C(0)$ is the spectral density function.

::: {.theorem #valid-covariogram}
Theorem For any normalized continuous positive-definite function $f$ on $G$ (normalization here means that $f$ is 1 at the unit of $G$ ), there exists a unique probability measure $\mu$ on $\widehat{G}$ such that
$$
f(g)=\int_{\widehat{G}} \xi(g) d \mu(\xi)
$$
:::

## Variogram Model Fitting

Considering variogram validity prerequisites, variogram estimators presented in section \@ref(estimator-of-variogram) can not be used directly for spatial prediction. The key idea of variogram model fitting is to search for a valid variogram that is closest to the data.

When there are more than one valid model, probably from method illustrated from section \@ref(maximum-likelihood-estimator) to \@ref(least-squares-esimator), **corss validation** is an important tool for model checking.

### EDA Check Before Model Fitting

- Check assumptions such as constant mean, constant marginal variance, stationarity, intrinsic stationarity;
- Check the Gaussian assumption;  
    - If the data is approximately normally distributed, modelling the first two moments are quite enough.
    - If some evidence of non-Gaussianity is found, the variogram estimator defined in section \@ref(estimator-of-variogram) is still valid, because they can be viewed as moment estimators which are independent of Gaussian assumption. 
- Isolate suspicious observations for further study;
- Choose a appropriate parametric variogram model.

From now on, we consider

### Maximum Likelihood Estimator

- Assumptions:
    - Suppose the data is generated from a stationary Gaussian process with a parametric variogram model $P=\{2 \gamma: 2 \gamma(h)=2 \gamma(h ; \theta) ; \theta \in \Theta\}$. Then the goal is to estimate $\theta$.
    - Suppose we also observed some covariates $X_1,...,X_n$ which are linearly related to $Z$.
- Model: $Z=\left(Z\left(s_{1}\right), \ldots, Z\left(s_{n}\right)\right) \sim N(X \beta, \Sigma(\theta))$ where $\Sigma(\theta)=\left(\operatorname{cov}\left(Z\left(s_{i}\right), Z\left(s_{j}\right)\right)\right)$.
- Goal: find parameter estimation $(\hat \beta, \hat \theta)$ to minimize loglikelihood function \@ref(eq:loglikelihood).
    - Where there is no spatial correlation, this amounts to a simple linear regression model. The solution is $\Sigma(\theta)=\sigma^{2} I$ and $\theta=\sigma^{2}.$
    - General solution: $\hat{\sigma}^{2}=\sum\left(Z\left(s_{i}\right)-X_{i} \hat{\beta}\right)^{2} / n$
        - $\hat{\sigma}^{2}$ is biased.
        - For $E\left(\hat{\sigma}^{2}\right)=(n-q) / n \sigma^{2}$, the larger $q$ is, the more biased the estimator gets. 

\begin{equation}
L(\beta, \theta)=\frac{n}{2} \log (2 \pi)+\frac{1}{2} \log |\Sigma(\theta)|+\frac{1}{2}(Z-X \beta)^{\prime} \Sigma(\theta)^{-1}(Z-X \beta)
(\#eq:loglikelihood)
\end{equation}
    
- Improvements: To de-bias $\hat \theta$, which is $\hat{\sigma}^{2}$ is out case, restricted maximum likelihood estimator (EMLE) is raised.
    - Estimator form: $\hat{\theta}=\int_{\beta} \exp (-L(\beta, \theta))(d \beta)$


Once variogram is estimated, then covariogram can be easily get by $2\gamma(h) = 2(C(0) - C(h))$

### MINQ Estimation

Applicable Situation: the variance matrix of the data is linear in its parameters, i.e. $\Sigma(\theta)=\theta_{1} \Sigma_{1}+\ldots, \theta_{m} \Sigma_{m}$. The MINQ approach is particularly appropriate for a variance components model.

#### Least Squares Esimator

The main idea of Least Squares Esimator (LSE) is the same as least square regression. The goal function to minimize is the square error.

\begin{equation}
\sum_{j=1}^{K}\left\{2 \gamma^{\#}(h(j) e)-2 \gamma(h(j) e ; \theta)\right\}^{2}
(\#eq:lse)
\end{equation}

If we not $2 \boldsymbol{\gamma}^{\#}=\left(2 \gamma^{\#}(h(1)), \ldots, 2 \gamma^{\#}(h(K))\right)$ with covariance matrix $\operatorname{var}\left(2 \boldsymbol{\gamma}^{\#}\right)=V$, the goal to minimize is tranformed into 

\begin{equation}
\left(2 \boldsymbol{\gamma}^{\#}-2 \boldsymbol{\gamma}(\theta)\right)^{\prime} V^{-1}\left(2 \boldsymbol{\gamma}^{\#}-2 \boldsymbol{\gamma}(\theta)\right)
(\#eq:lse2)
\end{equation}

- Notice that although a direction vector $e$ is included in formula \@ref(eq:lse), multiple directions could also be accounted for by adding the appropriate squared differences.
- Drawbacks: The least squares estimator takes no cognizance of the distributional variation and covariation of the generic estimator.
- Improvement:
    - Weighted Least Squares (WLS): Replace $V$ by a diagonal matrix $\left.\Delta=\operatorname{diag}\left(\operatorname{var}\left(2 \gamma^{\#}(h(1))\right), \ldots, 2 \gamma^{\#}(h(K))\right)\right)$.
    - Generalized Least Squares (GLS): use just the (asymptotic) second-order structure of the variogram estimator and does not make assumptions about the whole distribution of the data. When there is little knowledge of the true model, GLS is a good startingg choice.



















